% !TeX root = ../main.tex

\chapter{Experimental Setup}
\label{ch:05}

This chapter fixes the experimental ground rules shared by every result in
Chapter~\ref{ch:06}: the benchmark and its scoring semantics
(Section~\ref{sec:setup-realbench}), the models and serving infrastructure
(Section~\ref{sec:setup-models}), the metrics and the cost-accounting rules we
commit to (Section~\ref{sec:setup-metrics}), and a complete inventory of the
experimental runs together with their configurations
(Section~\ref{sec:setup-runs}). The inventory table doubles as the
reproducibility anchor of this thesis: every number reported later traces back
to exactly one row of Table~\ref{tab:run-inventory}.

\section{The RealBench Benchmark}
\label{sec:setup-realbench}

All end-to-end evaluations use RealBench~\citep{realbench2025}, a Verilog
generation benchmark built from three real, open-source IP projects rather
than from isolated textbook exercises. Each task provides a natural-language
specification of one target module, the source files of the modules it may
depend on, a golden reference implementation, and a self-checking testbench.
The generator must produce the target module; the surrounding project files
are supplied by the evaluation environment.

\begin{table}[htbp]
  \centering
  \caption[RealBench module-level task families]{The 60 module-level
    RealBench tasks used in this thesis, by source project.}
  \label{tab:families}
  \small
  \begin{tabular}{llrl}
    \toprule
    Family & Source project & Tasks & Character \\
    \midrule
    AES  & OpenCores AES-128 cipher core & 6  & self-contained datapath logic \\
    SDC  & OpenCores SD-card controller  & 14 & bus-facing control logic \\
    E203 & HummingBirdv2 E203 RISC-V core & 40 & deep hierarchy, reuse-rich \\
    \bottomrule
  \end{tabular}
\end{table}

The benchmark distinguishes \emph{module-level} tasks, which target one module
inside an otherwise-provided project, from \emph{system-level} tasks, which
ask for an entire subsystem. This thesis reports on the \textbf{60
module-level tasks} summarized in Table~\ref{tab:families}; system-level
tasks are excluded from all headline numbers because their specifications
(up to $\sim$704\,k characters for \texttt{e203\_cpu\_top}) exceed what any of
our serving configurations can process without aggressive condensation, and
because a 60-task universe with three distinct families is already large
enough to expose the phenomena this thesis studies. The three families span
a useful difficulty gradient: AES modules are small and algorithmically
self-contained, SDC modules implement Wishbone-facing control logic with
wide interfaces, and the E203 family covers a complete RISC-V processor
core whose modules instantiate one another --- the setting in which IP reuse
has the most to offer.

\paragraph{Scoring.}
RealBench scores with its own per-task \texttt{Makefile}, and its verdict is
deliberately strict. A sample is counted as a \emph{syntax} pass only if
Verilator~\citep{verilator} compiles the full project --- generated module,
provided dependencies, reference model, and testbench together --- with
\textbf{no \texttt{\%Error} and no \texttt{\%Warning}} outside the Makefile's
explicit waiver list; warnings are fatal. A sample is a \emph{functional}
pass (``pass'' for short) only if it also compiles and the testbench
simulation reports zero output mismatches against the golden reference.
Two consequences of this design recur throughout the thesis. First, our
syntax numbers are conservative relative to benchmarks that fail only on
errors (Section~\ref{sec:threats}). Second, the environment imposes two
non-negotiable structural contracts on generated code: a dependency that the
environment ships as a file must be \emph{instantiated, never re-declared}
(re-declaration is a fatal \texttt{MODDUP}), and the testbench's DUT
instantiation fixes the exact port list (a missing port is a fatal
\texttt{PINNOTFOUND}). Section~\ref{sec:pipeline-generator} describes how the
pipeline internalizes both contracts.

\section{Models and Serving Infrastructure}
\label{sec:setup-models}

All generation uses open-weight models served locally through
vLLM~\citep{kwon2023vllm} behind an OpenAI-compatible endpoint, on
university/NCHC GPU nodes reached through SSH port forwarding. Three models
appear in this thesis:

\begin{itemize}
  \item \textbf{gpt-oss-20B}~\citep{openai2025gptoss} --- the workhorse model.
    All ablations (grounding, repair cache, repair budgets, routing) run on
    this model, so their comparisons are model-controlled.
  \item \textbf{gpt-oss-120B}~\citep{openai2025gptoss} --- used to test which
    findings survive a $6\times$ scale increase
    (Sections~\ref{sec:results-failure} and~\ref{sec:results-routing}).
  \item \textbf{Qwen3-235B}~\citep{yang2025qwen3} --- used in supplementary
    repair-budget runs; it does not participate in headline comparisons.
\end{itemize}

Two serving quirks shaped the implementation and are documented here because
they are invisible in the results yet load-bearing for reproduction. First,
the deployment returns answers in the \texttt{reasoning\_content} field with
an empty \texttt{content} field whenever tool schemas are attached to a
request; the client transparently falls back to the reasoning field and, for
the planner, re-prompts without tools (Section~\ref{sec:pipeline-planner}).
Second, the server caps completion length rather than rejecting a request
when prompt and requested tokens together exceed the 131{,}072-token context
window, but returns HTTP~400 when the prompt alone exceeds it --- the failure
mode behind the ``not generated'' bucket of Section~\ref{sec:results-failure}.

\section{Metrics and Cost Accounting}
\label{sec:setup-metrics}

\paragraph{Quality.}
For a run with $n$ samples per task, we report the \emph{syntax rate} and
\emph{pass rate} as per-sample means over all generated records, and
\emph{pass@k} per task with the standard unbiased
estimator~\citep{chen2021humaneval}
\[
  \text{pass@}k \;=\; \mathbb{E}_{\text{tasks}}
  \left[ 1 - \binom{n-c}{k} \Big/ \binom{n}{k} \right],
\]
where $c$ is the number of passing samples for the task, averaged over the 60
tasks. We additionally report \emph{solved tasks}: the number of tasks with
at least one passing sample, which equals $60 \cdot \text{pass@}n$ and is the
quantity a practitioner who can afford $n$ attempts actually cares about.
Pass@k carries an independence assumption across a task's samples; one of
our own techniques (the run-scoped repair cache,
Section~\ref{sec:results-cache}) violates it by design, so all reported
pass@k values come from arms where samples are independent, and run-scoped
numbers are shown only to quantify that specific effect.

\paragraph{Cost.}
We use two cost measures. \emph{Estimated tokens} are computed from
prompt/response character counts with a calibrated 4.38-characters-per-token
heuristic and summed per run; \emph{compute time} ($\Sigma$wall) is the sum
of per-sample wall-clock seconds, which measures total machine work
independently of the harness's $\sim$80-way request concurrency. Three
accounting rules apply throughout, each forced by a measurement hazard we
identified in our own data: (i)~token totals from runs that were resumed
after interruption are not cited as costs, because resumed samples are
re-read from disk with zero logged tokens; (ii)~token totals for the direct
flow are lower bounds, because its repair-loop calls are not instrumented;
and (iii)~wall-clock is never compared \emph{across} runs executed on
different days, because the shared vLLM server's load varied --- only
within-run sums and per-solved-task ratios are reported. These hazards and
their blast radius are discussed together in Section~\ref{sec:threats}.

\section{Run Inventory}
\label{sec:setup-runs}

Table~\ref{tab:run-inventory} lists every experimental run this thesis
reports, the configuration that distinguishes it, and the section that
consumes it. Short labels (left column) are used throughout
Chapter~\ref{ch:06}. All runs share the same task discovery, per-task
catalog construction, and RealBench evaluation harness, so rows differ only
in the flow and the flagged configuration. Unless stated otherwise, sampled
runs use temperature~0.2 (the runner default) and the two-stage pipeline of
Chapter~\ref{ch:03}; ``$6/4$'' style entries abbreviate the syntax/functional
repair-attempt budgets of Section~\ref{sec:pipeline-repair}.

\begin{table}[htbp]
  \centering
  \caption[Inventory of experimental runs]{Inventory of all reported runs.
    Every number in Chapter~\ref{ch:06} traces to one row. Directory names
    are given in Table~\ref{tab:run-dirs}.}
  \label{tab:run-inventory}
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{llllp{0.34\textwidth}}
    \toprule
    Label & Model & $60\times n$ & Temp. & Distinguishing configuration \\
    \midrule
    \multicolumn{5}{l}{\emph{Baselines and headline pipeline}} \\
    \textsc{Direct-t0}   & 20B  & $\times 1$  & 0   & spec $\to$ model, no plan, no repair \\
    \textsc{Pipe-t0}     & 20B  & $\times 1$  & 0   & grounded planner + generator, cache off \\
    \midrule
    \multicolumn{5}{l}{\emph{Repair-cache trio (identical except cache scope)}} \\
    \textsc{Cache-off}   & 20B  & $\times 20$ & 0.2 & repair cache off \\
    \textsc{Cache-task}  & 20B  & $\times 20$ & 0.2 & repair cache, task scope \\
    \textsc{Cache-run}   & 20B  & $\times 20$ & 0.2 & repair cache, run scope \\
    \midrule
    \multicolumn{5}{l}{\emph{Model scale}} \\
    \textsc{Pipe-120b}   & 120B & $\times 10$ & 0.2 & as \textsc{Cache-run}, tool-calling probe \\
    \midrule
    \multicolumn{5}{l}{\emph{Functional repair and budgets}} \\
    \textsc{Func-2/4}    & 20B  & $\times 10$ & 0.2 & functional repair + spec slice, budgets $2/4$ \\
    \textsc{Func-10/10}  & 20B  & $\times 10$ & 0.2 & as above, budgets raised to $10/10$ \\
    \midrule
    \multicolumn{5}{l}{\emph{Routing (Chapter~\ref{ch:04} arms)}} \\
    \textsc{Hybrid}      & 20B  & $\times(10{+}10)$ & 0.2 & no router: both flows on every task \\
    \textsc{Router-t1}   & 20B  & $\times 10$ & 0.2 & plan-probe tier only, budgets $6/4$ \\
    \textsc{Router-2t}   & 20B  & $\times 10$ & 0.2 & full cascade (Tier-0 LLM + probe), $6/4$ \\
    \textsc{Router-2t-120b} & 120B & $\times 10$ & 0.2 & as \textsc{Router-2t} \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[htbp]
  \centering
  \caption[Run labels to repository directories]{Mapping from run labels to
    repository directories under \texttt{runs/}.}
  \label{tab:run-dirs}
  \footnotesize
  \begin{tabular}{ll}
    \toprule
    Label & Directory \\
    \midrule
    \textsc{Direct-t0}   & \texttt{realbench\_direct\_model} \\
    \textsc{Pipe-t0}     & \texttt{agentic\_plan\_legacy\_realbench\_plan\_hallu\_fix\_t0} \\
    \textsc{Cache-off}   & \texttt{agentic\_plan\_legacy\_realbench\_oss20b\_plan\_hallu\_no\_rep\_cache} \\
    \textsc{Cache-task}  & \texttt{agentic\_plan\_legacy\_realbench\_oss20b\_plan\_hallu\_task\_rep\_cache} \\
    \textsc{Cache-run}   & \texttt{agentic\_plan\_legacy\_realbench\_oss20b\_plan\_hallu} \\
    \textsc{Pipe-120b}   & \texttt{agentic\_plan\_legacy\_realbench\_oss120b\_plan\_hallu\_tool\_call} \\
    \textsc{Func-2/4}    & \texttt{repair\_spec\_slice\_oss20b\_func\_v2} \\
    \textsc{Func-10/10}  & \texttt{repair\_spec\_slice\_oss20b\_func\_v2\_time-err-off\_10syn\_10func} \\
    \textsc{Hybrid}      & \texttt{hybrid\_oss20b\_10syn\_10func\_rep\_spec} \\
    \textsc{Router-t1}   & \texttt{T1\_router\_oss20B\_sync6\_func4} \\
    \textsc{Router-2t}   & \texttt{Full-2T\_router\_oss20B\_sync6\_func4} \\
    \textsc{Router-2t-120b} & \texttt{Full-2T\_router\_oss120B\_sync6\_func4} \\
    \bottomrule
  \end{tabular}
\end{table}

Two rows deserve advance warning because they differ from their comparison
partners in more than one variable. \textsc{Func-10/10} changed both the
repair budgets ($2/4 \to 10/10$) and a scoring waiver
(\texttt{TIMESCALEMOD} warnings suppressed) relative to \textsc{Func-2/4};
Section~\ref{sec:results-budget} disentangles the two. And \textsc{Hybrid}
predates the direct-flow repair loop that the router runs include, besides
using the larger $10/10$ budgets; Section~\ref{sec:results-routing}
therefore treats \textsc{Router-t1} $\to$ \textsc{Router-2t} as the only
single-variable routing comparison and labels the comparison against
\textsc{Hybrid} as directional. One benchmark-side event also splits the
timeline: the RealBench AES S-box testbench was repaired upstream on
June~25,~2026 after we reported the race documented in
Section~\ref{sec:results-artifacts}; runs before and after that date are
never compared on the two affected tasks.
