% !TeX root = ../main.tex

\chapter{Background and Related Work}
\label{ch:02}

This chapter sets out the hardware-design background the thesis assumes
(Section~\ref{sec:bg-rtl}), then surveys the four research threads it
builds on: LLMs for Verilog generation
(Section~\ref{sec:rw-llm-verilog}), verification-in-the-loop repair
(Section~\ref{sec:rw-repair}), retrieval-augmented code generation
(Section~\ref{sec:rw-rag}), and LLM routing and cascades
(Section~\ref{sec:rw-routing}). Section~\ref{sec:rw-positioning} positions
this thesis against them.

\section{RTL Design, Verification, and IP Reuse}
\label{sec:bg-rtl}

Register-transfer-level (RTL) design expresses digital hardware as
synchronous transfers between registers through combinational logic,
written in a hardware description language --- here Verilog/SystemVerilog
\citep{ieee1800}. Unlike software, RTL code carries a hard structural
interface. A module's port list is a physical contract with its
environment, and simulation semantics (blocking vs.\ nonblocking
assignment, delta-cycle scheduling) can make two logically equivalent
descriptions observably different to a testbench. That subtlety
comes back in Section~\ref{sec:results-artifacts}. Correctness is
established by tools, not review. This thesis uses
Verilator~\citep{verilator} for lint and compiled simulation and
Yosys~\citep{wolf2013yosys} for synthesis-based structural analysis.

Industrial hardware productivity rests on \emph{IP reuse}, the practice
of assembling verified, pre-existing blocks rather than rewriting them,
codified two decades ago~\citep{keating2002reuse}. A working chip team
spends much of its effort not on new datapaths but on \emph{integration}
--- instantiating existing modules and wiring their interfaces correctly.
This observation is the design premise of Chapter~\ref{ch:03}. If
real-world RTL work is reuse-dominated, an RTL-generating agent should
plan reuse explicitly rather than synthesize every line from scratch. The
open-source projects our benchmark draws from --- an AES cipher core and
an SD-card controller from OpenCores~\citep{opencores} and the
HummingBirdv2 E203 RISC-V processor~\citep{e203hbirdv2} --- are themselves
products of this reuse culture.

\section{Large Language Models for Verilog Generation}
\label{sec:rw-llm-verilog}

\paragraph{Benchmarks.}
Evaluation infrastructure for LLM-generated Verilog began with
fine-tuning-era studies~\citep{pearce2020dave, thakur2023verigen} and
then settled into standard suites. VerilogEval~\citep{liu2023verilogeval}
offers machine-checked HDLBits-derived problems with the pass@k protocol
inherited from HumanEval~\citep{chen2021humaneval}, later revised with
specification-to-RTL tasks~\citep{pinckney2025verilogeval2}.
RTLLM~\citep{lu2024rtllm} adds larger open-ended designs judged on syntax,
functionality, and quality. These suites consist mostly of
\emph{self-contained} single modules. RTL-Repo~\citep{allam2024rtlrepo}
moved evaluation toward multi-file repository context.
RealBench~\citep{realbench2025}, used throughout this thesis
(Section~\ref{sec:setup-realbench}), builds tasks from complete, real IP
projects in which a target module must integrate against provided
dependencies and survive a strict warnings-as-fatal harness.
Table~\ref{tab:bench-stats} quantifies how far apart these settings are.
A RealBench module task averages an order of magnitude more code lines
and synthesized cells than a VerilogEval problem, and is specified by a
document two orders of magnitude longer. Alone among the suites, it
instantiates submodules at all. The distinction is not cosmetic.
Chapter~\ref{ch:06} shows that the phenomena that dominate RealBench
(interface contracts, reuse, integration) are precisely the ones
self-contained benchmarks cannot exhibit.

\begin{table}[htbp]
  \centering
  \caption[Design and specification scale across Verilog benchmarks]{Design
    and specification scale across Verilog generation benchmarks (per-design
    averages, reproduced from the RealBench study~\citep{realbench2025}).
    ``Cells'' =
    synthesized circuit cells; ``Doc.\ lines'' = specification length.}
  \label{tab:bench-stats}
  \small
  \setlength{\tabcolsep}{5pt}
  \begin{tabular}{lrrrrr}
    \toprule
    Benchmark & Designs & Code lines & Cells & Submodules & Doc.\ lines \\
    \midrule
    Thakur et al.~\citep{thakur2023verigen} & 17 & 16.6 & 7.9 & 0 & 1.8 \\
    RTLLM v1.1~\citep{lu2024rtllm} & 29 & 56.1 & 48.1 & 0.3 & 22.3 \\
    RTLLM v2~\citep{lu2024rtllm} & 50 & 46.1 & 33.7 & 0.2 & 20.8 \\
    VerilogEval-human~\citep{liu2023verilogeval} & 156 & 15.8 & 31.4 & 0 & 5.7 \\
    VerilogEval-machine~\citep{liu2023verilogeval} & 143 & 13.9 & 6.4 & 0 & 3.2 \\
    VerilogEval v2~\citep{pinckney2025verilogeval2} & 156 & 16.1 & 31.4 & 0 & 16.5 \\
    RealBench-module~\citep{realbench2025} & 60 & 241.2 & 520 & 1.6 & 197.3 \\
    RealBench-system~\citep{realbench2025} & 4 & 3537.3 & 1.1M & 14.5 & 2900 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Models and systems.}
One line of work improves the generator itself. VeriGen fine-tunes on
harvested Verilog corpora~\citep{thakur2023verigen}; ChipNeMo domain-adapts
LLMs across chip-design tasks~\citep{liu2023chipnemo};
RTLCoder~\citep{liu2024rtlcoder}, CodeV~\citep{zhao2024codev},
SiliconMind-V1~\citep{siliconmind2025}, and
CraftRTL~\citep{liu2025craftrtl} push open models past commercial
baselines with domain-specialized training pipelines --- CodeV through
multi-level code summarization, SiliconMind-V1 through multi-agent
distillation and debug-reasoning workflows, and CraftRTL by adding
correct-by-construction non-textual
representations and targeted repair data. Conversational studies such as Chip-Chat~\citep{blocklove2023chipchat}
probed interactive design with a human in the loop. This thesis is
orthogonal to all of them. We hold the base models fixed (off-the-shelf
open weights, Section~\ref{sec:setup-models}) and engineer the
\emph{system around} the model --- planning, grounding, verification,
repair, and routing. Better base models slot in unchanged; indeed
Section~\ref{sec:results-failure} measures exactly what a $6\times$ model
scale-up does and does not buy under a fixed system.

\section{Verification-in-the-Loop Generation and Repair}
\label{sec:rw-repair}

Feeding tool feedback back into generation is the field's standard answer
to the unreliability of one-shot code. In software,
self-repair with execution feedback is well studied and its limits
documented~\citep{olausson2024selfrepair}; agentic framings add explicit
reasoning-action loops~\citep{yao2023react}, verbal
self-reflection~\citep{shinn2023reflexion}, and purpose-built
agent-computer interfaces~\citep{yang2024sweagent}. In hardware,
AutoChip iterates generation against compiler and simulation
errors~\citep{thakur2024autochip}; RTLFixer targets syntax repair with
retrieval over an error-pattern database~\citep{tsai2024rtlfixer}; MEIC
frames RTL debug as an iterated LLM workflow~\citep{xu2024meic}; and
VerilogCoder couples an autonomous agent with graph-based planning and an
AST/waveform-tracing tool, reporting strong VerilogEval-Human
results~\citep{ho2025verilogcoder}.

The repair machinery of Section~\ref{sec:pipeline-repair} belongs to this
family but contributes three things the prior systems do not measure.
First, a \emph{semantic repair cache} that replays the model's own
verified diagnostic$\to$diff fixes across samples. It is related in spirit to
RTLFixer's error-pattern retrieval, but populated online from verified
successes rather than curated offline, and evaluated for what it changes
(sample density) and what it cannot (task solvability,
Section~\ref{sec:results-cache}). Second, an explicit
\emph{oracle-feedback accounting}. Our functional repair loop consumes the
scoring testbench's mismatch signal, so it is quarantined as a separate
arm, a methodological line most repair papers do not draw. Third, a
\emph{budget calibration} showing both repair phases are fix-fast-or-never
(Section~\ref{sec:results-budget}), turning repair-loop depth from
folklore into a measured engineering parameter.

\section{Retrieval-Augmented Code Generation}
\label{sec:rw-rag}

Retrieval-augmented generation grounds model output in external
sources~\citep{lewis2020rag}; for code, retrieval over repositories
supplies cross-file context that no prompt window holds natively, whether
by similar-code lookup~\citep{lu2022reacc} or iterative repository-level
retrieval interleaved with generation~\citep{zhang2023repocoder}. Our
per-task IP catalog (Section~\ref{sec:pipeline-overview}) is a
deliberately narrow instance. Retrieval is scoped to the task's own
dependency closure, never across tasks, and its product is not
only free-form context but a \emph{closed vocabulary} of instantiable
modules. The planner queries this catalog through its retrieval tools,
and the catalog digest is also injected into its prompt. Every
reuse decision in its output is grounded against the same vocabulary
post-hoc (Section~\ref{sec:results-grounding}), so retrieval serves as
constraint as well as context.

\section{LLM Routing and Cascades}
\label{sec:rw-routing}

A parallel literature allocates queries across models of different cost.
FrugalGPT sequences models with a learned stopping
rule~\citep{chen2023frugalgpt}; Hybrid LLM trains a router to send easy
queries to a small model~\citep{ding2024hybridllm}; RouteLLM learns
routers from human preference data~\citep{ong2024routellm}; AutoMix
self-verifies before escalating~\citep{aggarwal2024automix}. All of these
route between \emph{models} on general text tasks, assuming the larger
model weakly dominates in quality. Our router (Chapter~\ref{ch:04})
differs on both axes. It routes between \emph{workflows} over the same
model --- a cheap direct flow and an expensive agentic pipeline --- and
the expensive flow does \emph{not} dominate. On self-contained algorithmic
modules the pipeline is actively harmful
(Section~\ref{sec:router-motivation}), so routing is about matching task
\emph{character} to flow \emph{shape}, not buying quality with compute.
Its cascade structure is also domain-specific. The mid-tier probe
inspects a structured artifact the pipeline produces anyway (the plan),
so the escalation step costs one planner call rather than a full
duplicate generation.

\section{Positioning}
\label{sec:rw-positioning}

\begin{table}[htbp]
  \centering
  \caption[Positioning against related systems]{This thesis against the
    closest related systems. ``Reuse planning'' = explicit grounded
    IP-reuse decisions; ``workflow routing'' = per-task choice among
    generation flows.}
  \label{tab:positioning}
  \footnotesize
  \setlength{\tabcolsep}{3.5pt}
  \begin{tabular}{lccccc}
    \toprule
    System & Repair loop & Retrieval & Reuse planning & Workflow routing & Repo-realistic eval \\
    \midrule
    VeriGen~\citep{thakur2023verigen}       & --         & --     & --     & --     & -- \\
    AutoChip~\citep{thakur2024autochip}     & \checkmark & --     & --     & --     & -- \\
    RTLFixer~\citep{tsai2024rtlfixer}       & \checkmark & \checkmark & -- & --     & -- \\
    VerilogCoder~\citep{ho2025verilogcoder} & \checkmark & \checkmark & -- & --     & -- \\
    RepoCoder~\citep{zhang2023repocoder}    & --         & \checkmark & -- & --     & \checkmark \\
    RouteLLM~\citep{ong2024routellm}        & --         & --     & --     & model-level & -- \\
    \textbf{This thesis}                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
    \bottomrule
  \end{tabular}
\end{table}

To our knowledge, this thesis is the first system to combine
catalog-grounded IP-reuse \emph{planning} with per-task \emph{workflow
routing} for RTL generation, and the first to evaluate the combination on
a repository-realistic benchmark under controlled ablations
(Table~\ref{tab:positioning}). By design, it also contributes the
kind of controlled diagnostic measurements the surveyed literature
rarely reports: a grounding pass that makes every reuse decision in
every plan auditable, a repair cache whose effect is precisely located
(sample density inside solvable tasks), a budget calibration that pins
down the cost-efficient repair depth, and a golden-model audit
methodology that keeps the benchmark signal trustworthy
(Section~\ref{sec:results-artifacts}).
