% !TeX root = ../main.tex

\chapter{Cascade Routing between Direct and Pipeline Flows}
\label{ch:04}

Comparing the pipeline of Chapter~\ref{ch:03} with direct generation
task by task, we notice that the two implementations reach
\emph{different solution spaces}: each solves a set of tasks the other
does not. Diving into that solution variety shows a clean pattern.
Direct generation has the better ability on algorithm and FSM design
--- the model writes a self-contained datapath or state machine as one
coherent piece --- while pipeline-plan generation raises accuracy
substantially on integration and routing problems, where the work is
arranging and wiring existing modules. This division matches the purpose
of this thesis exactly: IP-reuse problems mostly need several existing
modules arranged and integrated correctly, which is precisely where the
plan pays, while the residual algorithmic tasks are best served by
letting the model generate directly. This chapter presents the second
contribution of the thesis: a per-task \emph{cascade router} that
predicts, before spending pipeline-scale compute, which of the two
flows a task should take --- the full agentic pipeline, or a cheap
\emph{direct} single-shot generation.
Section~\ref{sec:router-motivation} establishes why the two
flows are complementary; Section~\ref{sec:router-classes} analyzes the
two solved sets and defines the task classes and the oracle labels used
to evaluate routing; Section~\ref{sec:router-design} describes the
two-tier cascade; and Section~\ref{sec:router-metrics} defines the
routing-specific metrics used in Section~\ref{sec:results-routing}.

\section{Motivation: Two Complementary Flows}
\label{sec:router-motivation}

Running \emph{both} flows on all 60 tasks (the \textsc{Hybrid} arm of
Table~\ref{tab:run-inventory}) reveals a clean division of labor. With ten
samples per flow, the pipeline alone solves 20 tasks and the direct flow
alone solves 22 --- but their union solves 30, because the two sets overlap
only partially. Wrapper and integration modules pass under the pipeline,
which supplies the exact interfaces of the submodules they must instantiate
and wire; self-contained algorithmic modules pass under direct generation,
which lets the model write the datapath in one coherent piece. Neither
flow dominates: each one's strength is the other's blind spot.

\paragraph{Why planning trades away algorithm/FSM ability.}
That the pipeline gives up ground on algorithmic and FSM tasks is not an
implementation accident; it follows from what planning does. First,
decomposition fragments coherence: a state machine or an arithmetic
datapath is one tightly coupled artifact, and a plan that splits it into
submodule pieces forces the generator to implement fragments and glue
where a single \texttt{always}-block idiom --- exactly the pattern the
model knows best from pretraining --- would have expressed the whole
behavior. Second, the plan is a paraphrase: it restates the
specification's behavior in summarized form, and for algorithmic logic
the value lives in exact details (encodings, priorities, edge conditions)
that summarization inevitably coarsens; the generator then optimizes
toward the paraphrase rather than the source. Third, the plan's chief
asset --- verbatim submodule interfaces --- is worth little on a task
with nothing to instantiate, so the algorithmic task pays planning's
context and structure overhead without collecting its benefit. Routing
exists to give each task only the machinery it profits from.

The obvious exploitation of complementarity --- run both flows on every
task --- is also the most expensive one, and it wastes most of its budget:
every direct sample spent on a wrapper module and every pipeline run spent
on a self-contained ALU is compute that predictably converts to nothing.
A router that sends each task to its matching flow can, in principle, reach
the union coverage at roughly half the cost. The routing problem is to make
that prediction \emph{before} the expensive flow has been paid for.

\section{Task Classes and Oracle Labels}
\label{sec:router-classes}

Before formalizing classes, it is worth looking at the two solved sets
concretely rather than only through aggregate rates. In the
\textsc{Hybrid} run, eight tasks pass \emph{only} under the pipeline and
ten \emph{only} under direct generation
(Table~\ref{tab:flow-sets}). The membership alone is telling. The
pipeline-only set is wrappers, memories, and thin glue --- SRAM and
ITCM/DTCM wrappers (\texttt{e203\_srams}, \texttt{e203\_itcm\_ram},
\texttt{e203\_dtcm\_ram}), the register file and CSR-control shims
(\texttt{e203\_exu\_regfile}, \texttt{e203\_exu\_alu\_csrctrl},
\texttt{e203\_exu\_alu\_rglr}), the IFU integrator (\texttt{e203\_ifu}),
and a mini-decoder that instantiates the real decode logic
(\texttt{e203\_ifu\_minidec}). The direct-only set is algorithmic and
control-dense cores --- the AES cipher tops (\texttt{aes\_cipher\_top},
\texttt{aes\_inv\_cipher\_top}), instruction decode and branch
resolution (\texttt{e203\_exu\_decode}, \texttt{e203\_exu\_branchslv}),
address generation (\texttt{e203\_exu\_alu\_lsuagu}), and the SD-card
command/data engines and FIFOs (\texttt{sd\_cmd\_master},
\texttt{sd\_cmd\_serial\_host}, \texttt{sd\_rx\_fifo},
\texttt{sd\_tx\_fifo}).

\begin{table}[htbp]
  \centering
  \caption[Pipeline-only vs.\ direct-only solved sets]{The two
    flow-exclusive solved sets of the \textsc{Hybrid} run
    ($60\times(10{+}10)$), with structural statistics of their golden
    implementations (median / mean / range). ``Own cells'' = synthesized
    cells of the module's own logic after subtracting submodule
    instances.}
  \label{tab:flow-sets}
  \small
  \setlength{\tabcolsep}{5pt}
  \begin{tabular}{lll}
    \toprule
     & Pipeline-only (8 tasks) & Direct-only (10 tasks) \\
    \midrule
    Members &
      \begin{tabular}[t]{@{}l@{}}
        \texttt{e203\_srams}, \texttt{e203\_itcm\_ram},\\
        \texttt{e203\_dtcm\_ram}, \texttt{e203\_exu\_regfile},\\
        \texttt{e203\_exu\_alu\_csrctrl},\\
        \texttt{e203\_exu\_alu\_rglr}, \texttt{e203\_ifu},\\
        \texttt{e203\_ifu\_minidec}
      \end{tabular} &
      \begin{tabular}[t]{@{}l@{}}
        \texttt{aes\_cipher\_top}, \texttt{aes\_inv\_cipher\_top},\\
        \texttt{e203\_exu\_decode}, \texttt{e203\_exu\_branchslv},\\
        \texttt{e203\_exu\_alu\_lsuagu}, \texttt{e203\_ifu\_ift2icb},\\
        \texttt{sd\_cmd\_master}, \texttt{sd\_cmd\_serial\_host},\\
        \texttt{sd\_rx\_fifo}, \texttt{sd\_tx\_fifo}
      \end{tabular} \\
    \midrule
    Code lines        & 62 / 68 / 36--156          & 210 / 294 / 51--895 \\
    Spec size (chars) & 5.4k / 6.8k / 3.6k--15k    & 11.7k / 15.6k / 5k--35k \\
    Own cells         & 0 / 532 / 0--4063          & 818 / 1434 / 330--6275 \\
    Passing yield     & stable (mean 8.1/10)        & exploratory (mean 3.1/10) \\
    \bottomrule
  \end{tabular}
\end{table}

The structural statistics in Table~\ref{tab:flow-sets} separate the two
sets cleanly and make the eventual classifier concrete. Pipeline-only
modules are short (median 62 golden lines), briefly specified, and ---
the decisive feature --- synthesize to \emph{zero own cells} at the
median: their logic \emph{is} their submodule wiring. Direct-only
modules are $3$--$4\times$ longer, specified at twice the length, and
carry substantial own logic (median 818 cells) with no dominant
submodule structure. The two sets also differ in how they pass:
pipeline-solved tasks pass stably (8.1 of 10 samples on average, the
interfaces being supplied rather than guessed), while direct-solved
tasks pass by exploring the solution space (3.1 of 10, often a single
sample). These are exactly the observable characteristics --- delegation
vs.\ own logic, interface-dominated vs.\ behavior-dominated
specification --- that the router's feature extractor
(Section~\ref{sec:router-design}) is built to read from the
specification alone.

We formalize the division as two task classes:

\begin{description}
  \item[Class A (structural).] Wrapper, integration, and memory-encapsulation
    modules: the module's own logic is thin; its difficulty is wiring
    externally defined submodule interfaces correctly. Class~A tasks want
    the \textbf{pipeline}.
  \item[Class B (algorithmic).] Self-contained datapath or control modules
    --- cipher rounds, CRC, instruction decode, address generation: the
    difficulty is the logic itself, not integration. Class~B tasks want
    \textbf{direct} generation.
\end{description}

For evaluation only, each task receives an oracle label computed from its
\emph{golden} implementation: the reference RTL is synthesized with Yosys,
submodule instances are subtracted, and the task is labeled~A if its own
synthesized logic falls below a 250-cell threshold (or it is a recognized
regular-array wrapper), otherwise~B. The label also fixes the task's
\emph{target flow} (A$\to$pipeline, B$\to$direct). Oracle labels use golden
RTL that no generation-time component may see; they exist so that routing
accuracy can be scored, and they define the ceiling arm (\texttt{oracle}) in
the ablation of Section~\ref{sec:results-routing}.

\section{The Two-Tier Cascade}
\label{sec:router-design}

\begin{figure}[htbp]
  \centering
  \begin{tikzpicture}[
      font=\small,
      node distance=9mm and 14mm,
      box/.style={draw, rounded corners=2pt, align=center, inner sep=5pt,
                  minimum height=8mm},
      dec/.style={draw, diamond, aspect=2.2, align=center, inner sep=1.5pt},
      flow/.style={draw, rounded corners=2pt, align=center, inner sep=5pt,
                   minimum height=8mm, very thick},
      arr/.style={-{Stealth[length=2.5mm]}},
    ]
    \node[box] (spec) {task specification};
    \node[box, below=of spec] (feat) {Tier-0: feature extraction\\(LLM judge over the spec)};
    \node[dec, below=of feat] (pre) {pre-route\\decision};
    \node[flow, below left=13mm and 16mm of pre] (direct) {\textbf{direct flow}\\single-shot generation\\+ repair loops};
    \node[flow, below right=13mm and 16mm of pre] (pipe) {\textbf{pipeline flow}\\plan $\to$ generate\\+ repair loops};
    \node[box, below=17mm of pre] (plan) {Tier-1: run planner once,\\probe the structured plan};
    \node[dec, below=of plan] (probe) {plan\\verdict};
    \draw[arr] (spec) -- (feat);
    \draw[arr] (feat) -- (pre);
    \draw[arr] (pre) -| node[above, pos=0.25]{algorithmic (B)} (direct);
    \draw[arr] (pre) -| node[above, pos=0.25]{structural (A)} (pipe);
    \draw[arr] (pre) -- node[right]{uncertain} (plan);
    \draw[arr] (plan) -- (probe);
    \draw[arr] (probe) -| node[above, pos=0.25]{self-contained} (direct);
    \draw[arr] (probe) -| node[above, pos=0.25]{delegates (keep plan)} (pipe);
  \end{tikzpicture}
  \caption[Cascade router decision flow]{The two-tier cascade router. Tier-0
    routes confident tasks from spec-level features alone; only uncertain
    tasks pay for a planner call, whose structured plan Tier-1 probes. A
    task routed to direct after planning discards the plan (a ``wasted
    plan'').}
  \label{fig:router-flow}
\end{figure}

The router is a cascade of two increasingly expensive tests
(Figure~\ref{fig:router-flow}); a task exits at the first tier that is
confident.

\paragraph{Tier-0: spec pre-route.}
A feature extractor reads the specification and emits seven structured
judgments: whether the module delegates to submodules, is a memory
encapsulation, or is thin pass-through control (all class-A signals);
whether it contains a finite-state machine, clock-domain crossing, or a
datapath algorithm (class-B signals); and an estimate of the state bits the
module itself must hold, plus a confidence value. The production extractor
is a small LLM call that returns this JSON verdict (a regex-based keyword
extractor over the same features serves as the ablation baseline ---
Section~\ref{sec:results-routing}); the decision rule is deliberately
transparent: any algorithmic signal routes to \emph{direct}, else any
structural signal routes to \emph{pipeline}, else --- or when the
extractor's confidence is below threshold --- the task falls through as
\emph{uncertain}.

\paragraph{Why an LLM extractor rather than parsing.}
The features themselves are simple enough that keyword matching looks
tempting, but specification language is \emph{semantically misleading}
at exactly the decision boundary. A module can sound algorithmic while
being structural: \texttt{e203\_ifu\_minidec} is described as
performing ``preliminary decoding,'' yet it is a thin wrapper that
instantiates the real decoder --- a keyword extractor reads
``decoding'' and routes it to direct, where it has no access to the
decoder interface it must wire. The reverse confusion also occurs:
specifications for genuinely algorithmic FIFOs and CRC engines spend
most of their text on bus-facing port tables, so surface counting of
interface vocabulary misreads them as structural. Deciding the class
requires reading what the module \emph{does} across the whole document
--- whether the described behavior is delegated or computed --- which is
a semantic judgment, not a lexical one. The whole-spec LLM read resolves
both directions, and the ablation of Section~\ref{sec:results-routing}
quantifies the gap to the keyword baseline.

\paragraph{What falls through as uncertain.}
The uncertain residue is not noise; it is a coherent third type of
question. In the full-cascade run, exactly five tasks fell through:
\texttt{sd\_bd} (a buffer-descriptor manager), \texttt{sd\_data\_serial\_host}
(the SD data-line serial engine), \texttt{e203\_biu} (the bus interface
unit), \texttt{e203\_exu\_alu\_lsuagu} (the load/store address-generation
unit sharing the ALU datapath), and \texttt{e203\_exu\_wbck} (the
write-back arbiter). Their common shape: \emph{interface-arbitration
modules} that sit between protocols --- they neither delegate to
submodules (no class-A signal fires) nor implement a nameable algorithm
or FSM (no class-B signal fires), but instead mediate, arbitrate, and
translate between interfaces with modest own state. For every one of
the five, the extractor returned all-false features at full confidence:
the specification genuinely does not commit the module to either class,
and only the plan --- which makes the delegation structure explicit ---
can. These are precisely the tasks the second tier exists for.

\paragraph{Tier-1: plan probe.}
Uncertain tasks run the planner once --- the cheapest informative pipeline
component --- and the router probes the resulting structured plan's own
natural-language requirement summary. The plan is a good routing feature
for the same reason the condensation stage renders two views
(Section~\ref{sec:pipeline-planner}): the \emph{planner view} is built
to surface each module's delegation and reuse structure while the
\emph{generation view} carries its full behavioral requirements, so by
the time a plan exists, characteristics that were ambiguous in the raw
specification --- does this module instantiate the work or perform it?
--- have been made explicit in the plan's own vocabulary of reuse
decisions and requirement summaries. The probe simply reads that
vocabulary back. Decisive delegation phrasing
(``wraps'', ``encapsulates'', ``reuses existing'', or instantiation language
without any own-implementation verb) keeps the task in the pipeline, plan
in hand; clear self-contained-algorithm phrasing routes it to direct,
\emph{discarding} the plan. The asymmetry of the default is intentional:
the plan is already a sunk cost, so the probe bails to direct only when the
module is clearly self-contained. A plan generated and then discarded is
counted as a \emph{wasted plan} --- the cascade's overhead metric.

\paragraph{The direct flow.}
The direct arm is not a bare completion: it shares the generator's repair
machinery. A directly generated candidate that fails the internal lint
enters the same syntax repair loop, and (in arms where functional repair is
enabled) the same functional repair loop, with the plan-dependent prompt
sections simply absent. This matters for the routing economics: direct
generation costs roughly 10--12\,k tokens per sample against the pipeline's
90--153\,k, so every correctly-routed class-B task buys nearly an order of
magnitude of compute reduction at equal or better quality.

Figure~\ref{fig:router-samples} traces how the full-cascade run's
$60\times10 = 600$ samples actually move through these stages, together
with the volume the repair loops handle in each flow. Tier-0 routes 55
of 60 tasks from the specification alone (32 tasks to direct, 23 to the
pipeline); the remaining 5 tasks' samples pay one planner call each,
after which the probe keeps 28 samples in the pipeline and releases 22
to direct. Of the resulting 342 direct and 258 pipeline samples, the
syntax repair loop delivers 260 and 176 compiling candidates
respectively --- the direct flow compiles at 76.0\,\% with most samples
clean on arrival, while the pipeline flow leans on the loop harder, as
befits the giant integrators routed to it --- and the functional stage
converts 184 samples into passes across both flows.

\begin{figure}[htbp]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tikzpicture}[
      font=\small,
      box/.style={draw, rounded corners=2pt, align=center, inner sep=4pt,
                  minimum height=9mm},
      stage/.style={draw, rounded corners=2pt, align=center, inner sep=4pt,
                    minimum height=9mm, very thick},
      arr/.style={-{Stealth[length=2.5mm]}},
    ]
    \node[stage] (t0) {Tier-0\\spec pre-route};
    \node[box, left=8mm of t0] (all) {60 tasks\\$\times$10 = 600};
    \node[box, above right=6mm and 22mm of t0] (dir) {\textbf{direct flow}: 342\\(320 pre-routed\\+ 22 probe-routed)};
    \node[box, below right=6mm and 22mm of t0] (pipe) {\textbf{pipeline flow}: 258\\(230 pre-routed\\+ 28 probe-routed)};
    \node[stage, below=17mm of t0] (t1) {Tier-1 plan probe\\5 tasks (50 samples)};
    \node[box, right=9mm of dir] (dsyn) {syntax loop ($\leq$6)\\260 compile / 82 exh.};
    \node[box, right=9mm of pipe] (psyn) {syntax loop ($\leq$6)\\176 compile / 82 exh.};
    \node[box, right=9mm of dsyn] (dfun) {functional ($\leq$4)\\103 pass};
    \node[box, right=9mm of psyn] (pfun) {functional ($\leq$4)\\81 pass};
    \node[box, right=9mm of dfun, yshift=-13.5mm] (out) {184 passing\\34/60 solved};
    \draw[arr] (all) -- (t0);
    \draw[arr] (t0.north) |- node[left, font=\footnotesize, pos=0.12]{32 tasks} (dir.west);
    \draw[arr] (t0.south) |- node[left, font=\footnotesize, pos=0.12]{23 tasks} (pipe.west);
    \draw[arr] (t0) -- node[right, font=\footnotesize]{uncertain} (t1);
    \draw[arr] (t1.east) -- ++(6mm,0) |- node[right, font=\footnotesize, pos=0.3]{discard plan} ([yshift=-3mm]dir.west);
    \draw[arr] (t1.east) -- ++(6mm,0) |- node[right, font=\footnotesize, pos=0.35]{keep plan} ([yshift=-3mm]pipe.west);
    \draw[arr] (dir) -- (dsyn);
    \draw[arr] (pipe) -- (psyn);
    \draw[arr] (dsyn) -- (dfun);
    \draw[arr] (psyn) -- (pfun);
    \draw[arr] (dfun.east) -| ([xshift=4mm]dfun.east) |- ([yshift=2.5mm]out.west);
    \draw[arr] (pfun.east) -| ([xshift=4mm]pfun.east) |- ([yshift=-2.5mm]out.west);
  \end{tikzpicture}}
  \caption[Sample flow through the cascade]{How the full-cascade run's
    600 samples route through the tiers and the shared repair loops
    (gpt-oss-20B, budgets $6/4$). Tier-0 decides 55 of 60 tasks from the
    specification; the plan probe adjudicates the 5 uncertain tasks
    per-sample. Both flows run the same syntax and functional repair
    loops; counts give each stage's outcomes.}
  \label{fig:router-samples}
\end{figure}

\section{Routing Metrics}
\label{sec:router-metrics}

Section~\ref{sec:results-routing} evaluates routing arms on four axes:
\emph{coverage} (solved tasks and pass@k, against the \textsc{Hybrid} union
as the attainable ceiling); \emph{routing accuracy} (the confusion between
oracle class and chosen flow, where both misroute directions ---
A$\to$direct and B$\to$pipeline --- forfeit a would-be pass);
\emph{overhead} (wasted plans); and \emph{efficiency} (estimated tokens and
summed compute per solved task). The design goal is stated once and
measured throughout: reach the union coverage of both flows at the lowest
compute, with misroutes near zero.
