% !TeX root = ../main.tex

\chapter{Discussion, Limitations, and Conclusion}
\label{ch:07}

\section{Discussion}
\label{sec:discussion}

\paragraph{What actually moved the needle.}
Ranked by measured effect, the levers were: (1)~\emph{environment-faithful
verification and repair} --- compiling against the real file set,
injecting the testbench port contract, and condensing giant specs took the
pipeline from low-single-digit to ${\sim}18$--25\,\% pass, the single
largest step in the project's history; (2)~\emph{routing} --- matching
task character to flow shape held coverage while \emph{cutting} compute
by 39\,\% and per-solved-task tokens by roughly 40\,\%
(Section~\ref{sec:results-routing});
(3)~\emph{model scale} --- $6\times$ parameters bought $+9$\,pp syntax and
$+4$\,pp pass@1 with the system held fixed; (4)~\emph{repair caching} ---
$+5$--$7$\,pp syntax, density only. Smaller on the headline metric but
valuable elsewhere: grounding ($+0.6$\,pp, while buying auditability and
eliminating a failure class) and the functional repair loop ($+7$ net
passes, best run at a compact budget). The general lesson we draw is
unglamorous: in agentic code generation for strict environments,
\emph{fidelity between the inner loop and the scorer} is worth more
than additional intelligence in any single stage.

\paragraph{Grounding as a harness property.}
The planner's tool suite gives it evidence on demand --- catalog search,
candidate inspection, interface checks --- but the design never lets
plan correctness \emph{depend} on any particular trajectory through
those tools: the catalog vocabulary is also injected into the prompt,
the output schema is constrained to that vocabulary, and every emitted
reuse decision is resolved against the catalog post-hoc. The result is
that plans are correct by construction regardless of how the model
arrived at them, and the grounding audit
(Section~\ref{sec:results-grounding}) can verify it decision by
decision. We suspect this pattern --- \emph{make correctness a property
of the harness, layered over rather than delegated to the model's
tool-use behavior} --- transfers well beyond RTL.

\paragraph{Why routing pays here.}
The routing literature mostly buys quality with compute along one axis
(small model vs.\ large model). Our setting is different in kind: the
expensive flow's competence subsumes the cheap flow's, but the two
differ by an order of magnitude in cost per sample, and the scaffold
that makes the expensive flow strong on integration-heavy instances is
measurably counterproductive on self-contained ones
(Section~\ref{sec:router-motivation}) --- so routing to the cheap path
loses nothing there and saves nearly everything. We expect this shape
to recur wherever an elaborate agentic scaffold coexists with a cheap
direct path, because the routing signal (does this task's difficulty
live in its interfaces or in its logic?) is legible from the
specification itself. The corollary from our audit is equally general:
when a cheap flow suddenly looks \emph{more able} than the scaffold ---
rather than merely cheaper --- that is a claim worth auditing before it
is banked (Section~\ref{sec:results-artifacts}).

\paragraph{The functional ceiling.}
The binding constraint at the end of this work is unambiguous:
${\sim}60$--69\,\% of everything that compiles fails its testbench, a
share that neither model scale, nor caching, nor ten rounds of
oracle-fed repair meaningfully dents. Verilator diagnostics give the
repair loop a precise, local, actionable signal for syntax; a mismatch
count gives almost nothing --- and Section~\ref{sec:results-artifacts}
showed it can even be actively misleading. Progress on this tier will
require a fundamentally better feedback channel (waveform-level
localization, per-submodule benches, assertion synthesis), not more
attempts on the current one.

\section{Threats to Validity}
\label{sec:threats}

We collect here, in one place, every caveat that qualifies the numbers in
this thesis; each was also flagged inline where it bites.

\begin{enumerate}
  \item \textbf{Single-seed runs.} Temperature-0 arms are one sample per
    task; each sampled run is one seed. Cross-run deltas under
    ${\sim}2$--3\,pp are compatible with sampling noise. The claims we
    rest weight on (headline $3.0\times$; routing $-39\,\%$ compute at
    matched coverage; the flat pass@20) are far outside that band ---
    the routing arms' one-task set differences are inside it and are
    treated as noise.
  \item \textbf{Cross-run configuration drift.} Only three comparisons
    are single-variable: the cache trio, the temp-0 head-to-head, and
    \textsc{Router-t1}$\to$\textsc{Router-2t}. Everything else ---
    notably the router-vs-\textsc{Hybrid} comparison (repair budgets
    \emph{and} direct-flow repair differ) and the \textsc{Func-2/4}
    $\to$\textsc{Func-10/10} pair (a scoring waiver changed) --- is
    labeled directional where it appears.
  \item \textbf{Cost accounting.} Token totals from resumed runs
    undercount (resumed samples log zero tokens): the only clean
    full-run token cost is \textsc{Cache-off}'s 46.3\,M. Direct-flow
    repair calls are not instrumented, so direct-flow token totals ---
    including \textsc{Router-2t}'s 28.3\,M --- are lower bounds; the
    per-solved-task advantage in Table~\ref{tab:routing} (1.35\,M vs.\
    the base arm's measured 2.17\,M) would shrink somewhat, not
    reverse, under full instrumentation.
    Wall-clock is never compared across days (shared server, varying
    load).
  \item \textbf{pass@k independence.} The run-scoped cache couples a
    task's samples; all load-bearing pass@k values come from cache-off
    or task-scoped arms.
  \item \textbf{Scoring strictness.} The harness fails on warnings, so
    our syntax rates are conservative relative to error-only benchmarks
    --- and Section~\ref{sec:results-artifacts} showed our own internal
    gate was at times \emph{stricter than the scorer}, deflating four
    tasks.
  \item \textbf{Benchmark fix boundary.} The AES S-box testbenches were
    repaired upstream on June~25,~2026, after our report. Every run in
    this thesis predates the fix except where stated; no comparison
    spans the boundary on \texttt{aes\_sbox}/\texttt{aes\_inv\_sbox}.
  \item \textbf{Ref-wrap exclusion.} All rates exclude samples that
    instantiate a leaked \texttt{ref\_*} reference model
    (Section~\ref{sec:results-artifacts}). Detection is a syntactic
    screen over generated RTL; it flagged only unambiguous
    self-wraps and cannot, by construction, catch subtler uses of the
    leaked \emph{name} --- though the reference \emph{source} was never
    visible to the models, bounding what a subtler exploit could copy.
    The direct-flow prompt retains the leak in the archived runs; the
    definitive control is a re-run with the filtered prompt
    (Section~\ref{sec:future}).
  \item \textbf{Oracle feedback.} The functional repair loop consumes
    the scoring testbench's mismatch report. Its passes mean
    ``specification plus oracle feedback'' and are quarantined in
    Section~\ref{sec:results-funcrepair}; no headline number includes
    them. Relatedly, the internal lint compiles the testbench, so
    Verilator diagnostics can quote testbench lines into repair prompts;
    a flag exists to disable this for leak-free claims, at a syntax-rate
    cost we did not pay here.
  \item \textbf{Benchmark scope.} All results are on one benchmark's 60
    module-level tasks and three IP projects, with open-weight models of
    one family for all controlled comparisons. The family-level
    heterogeneity we observe (AES vs.\ E203) is itself evidence that
    results will vary across codebases.
\end{enumerate}

\section{Future Work}
\label{sec:future}

Five directions follow directly from the measured constraints.
\emph{First}, functional feedback without oracle leakage: synthesize
self-checking benches or assertions from the specification, so the
functional repair loop gets a mismatch signal the scorer never sees ---
and gate it off pure-combinational modules, where
Section~\ref{sec:results-artifacts} showed the signal can be corrupted.
\emph{Second}, combining the pipeline with a testbench-generation
pipeline to unlock \emph{fine-grained module-decomposition IP reuse}:
today the plan's decomposition stops at module granularity because only
the top module has an oracle, but a companion pipeline that generates a
bench per planned submodule would let the system decompose more
aggressively, verify each decomposed piece independently, and reuse IP
at the sub-block level --- turning the planner's decomposition from a
prompt-structuring device into a verified construction process.
\emph{Third}, interface-contract enforcement as a hard gate: the
testbench DUT instantiation is already injected as text; promoting it to
a mechanical post-generation port-diff check would delete the
\texttt{PINNOTFOUND} class that dominates giant-module syntax failures,
and the internal repair gate should mirror the scorer's warning waivers
exactly. \emph{Fourth}, selective decomposition for the giants:
RealBench's system-level tasks decompose into submodules that are
themselves benched module tasks, so a tree-structured generator could
verify each node against a \emph{real} testbench and compose upward ---
decomposition with per-node oracles, reserved for the scale-limited tier
where it pays, and the natural first customer of the
testbench-generation direction above. \emph{Fifth}, hardening the
router: the dominant misroute (A$\to$direct) argues for widening the
uncertainty band that triggers the plan probe; a replication with the
\texttt{ref\_*}-filtered direct prompt is the definitive control for
the ref-wrap exclusion of Section~\ref{sec:results-artifacts}; and
multi-seed replication would move the routing result from strongly
suggestive to statistically settled.

\section{Conclusion}
\label{sec:conclusion}

This thesis asked how far careful \emph{system} engineering --- planning,
grounding, verification, repair, and routing around fixed open-weight
models --- can push LLM-based RTL generation on a realistic,
reuse-dominated benchmark. The answer is: substantially, and measurably.
A two-stage pipeline that plans IP reuse against a grounded per-task
catalog and generates under environment-faithful verification reaches
$2.5\times$ the syntax rate and $3.0\times$ the pass rate of the same
model prompted directly, with the entire gain concentrated where reuse
exists to be exploited. A two-tier cascade router that matches task
character to generation flow holds that coverage --- matching its
pipeline-only base arm task for task --- at 39\,\% less compute and
the lowest token cost per solved task of any arm measured, less than a
third of the brute-force deployment's.

The thesis's second kind of contribution is measurement discipline over
the same system: grounding made every plan correct and auditable;
caching concentrated passing density inside solvable tasks; the repair
budgets were calibrated to where the rescues actually happen, recovering
the large majority of the functional phase's compute; and a golden-model
audit of persistent zeros cleared the models on two tasks, aligned our
own repair gate with the scorer on four, confirmed the benchmark
sound on the rest --- and, turned on our own passes, caught a
prompt-side leak that had inflated the routing ablation and corrected
it before it could mislead. These findings redirected the engineering
each time, and one was adopted upstream by the benchmark itself. The frontier, at the end, is not
syntax, not plans, and not scale: it is functional correctness, and the
next system to move it will be the one that finds its models a better
feedback signal than a mismatch count.
