% !TeX root = ../main.tex

\chapter{Discussion, Limitations, and Conclusion}
\label{ch:06}

\section{Discussion}
\label{sec:discussion}

\paragraph{What actually moved the needle.}
Ranked by measured effect, the levers were as follows. (1)~\emph{environment-faithful
verification and repair} --- compiling against the real file set,
injecting the testbench port contract, and condensing giant specs took the
pipeline from low-single-digit to ${\sim}18$--25\,\% pass, the single
largest step in the project's history; (2)~\emph{model scale} ---
$6\times$ parameters bought $+9$\,pp syntax and $+4$\,pp pass@1 with the
system held fixed; (3)~\emph{routing} --- matching task character to
flow shape held coverage at 21 of 60, level with the fitted rule it
replaces, with better per-sample
pass and syntax rates and no exposure to the benchmark's leaked
reference models; the original rule's apparent compute cut did
not survive the benchmark-gaming audit
(Section~\ref{sec:results-routing}); (4)~\emph{repair caching} ---
$+5$--$7$\,pp syntax, density only. Two levers were smaller on the
headline metric but valuable elsewhere --- grounding ($+0.6$\,pp, while
buying auditability and eliminating a failure class) and the functional
repair loop ($+7$ net passes, best run at a compact budget). The
general lesson is plain. In agentic code generation for strict
environments, \emph{fidelity between the inner loop and the scorer} is
worth more than additional intelligence in any single stage.

\paragraph{Grounding as a harness property.}
The planner's tool suite gives it evidence on demand --- catalog search,
candidate inspection, interface checks --- but the design never lets
plan correctness \emph{depend} on any particular trajectory through
those tools. The catalog vocabulary is also injected into the prompt,
the output schema is constrained to that vocabulary, and every emitted
reuse decision is checked against the catalog afterwards. The result is
that plans are correct by construction regardless of how the model
arrived at them, and the grounding audit
(Section~\ref{sec:results-grounding}) can verify it decision by
decision. This pattern --- \emph{make correctness a property
of the harness, layered over rather than delegated to the model's
tool-use behavior} --- should transfer well beyond RTL.

\paragraph{What routing buys here.}
The routing literature mostly buys quality with compute along one axis
(small model vs.\ large model). Our setting is different in kind. The
expensive flow can do everything the cheap flow can, but the two
differ by an order of magnitude in cost per sample, and the scaffold
that makes the expensive flow strong on integration-heavy tasks is
measurably counterproductive on self-contained ones
(Section~\ref{sec:router-motivation}). The corrected result shrinks
the payoff, not the direction. After the audit, the set the cheap flow
genuinely handles is narrow and also happens to be planning-cheap, so
at 20B routing buys per-sample quality and exposure control rather
than compute; the compute lever returns only where the cheap-safe set
widens into planning-expensive territory, and even the first candidate
at 120B turns out to be reachable by the deep-budget planning flow as well
(Section~\ref{sec:results-routing}). The setting's shape will recur
wherever an elaborate agentic scaffold coexists with a cheap direct
path, because the routing signal (does this task's difficulty live in
its interfaces or in its logic?) can be read off the specification
itself. Our audit adds a more general lesson. When a cheap flow
suddenly looks \emph{more able} than the scaffold, rather than merely
cheaper, audit that claim before believing it --- and before fitting
any rule to it (Section~\ref{sec:results-artifacts}).

\paragraph{The functional ceiling.}
The binding constraint at the end of this work is clear.
${\sim}60$--69\,\% of everything that compiles fails its testbench, a
share that model scale, caching, and ten rounds of
oracle-fed repair barely move. Verilator diagnostics give the
repair loop a precise, local, actionable signal for syntax; a mismatch
count gives almost nothing --- and Section~\ref{sec:results-artifacts}
showed it can even be actively misleading. Progress on this tier will
require a fundamentally better feedback channel (waveform-level
localization, per-submodule benches, assertion synthesis), not more
attempts on the current one.

\section{Threats to Validity}
\label{sec:threats}

This section collects every caveat that qualifies the numbers in
this thesis; each was also flagged inline where it applies.

\begin{enumerate}
  \item \textbf{Single-seed runs.} Temperature-0 arms are one sample per
    task; each sampled run is one seed. Cross-run deltas under
    ${\sim}2$--3\,pp are compatible with sampling noise. The claims we
    rest weight on (headline $3.0\times$; the flat pass@20;
    \textsc{Router-v2}'s per-sample pass@1 and syntax margins over the
    v1 rule, measured on 600 samples each) are far outside that band ---
    the routing arms' one-task set differences are inside it and are
    treated as noise.
  \item \textbf{Cross-run configuration drift.} Only four comparisons
    are single-variable: the cache trio, the temp-0 head-to-head,
    \textsc{Router-v1}$\to$\textsc{Router-v2} (decision rule only), and
    \textsc{Func-10/10}$\to$\textsc{Func-10/10-120b} (model scale
    only, though the two ran on different serving deployments, so only
    quality metrics --- not wall-clock --- are compared).
    Everything else ---
    notably the router-vs-\textsc{Hybrid} comparison (repair budgets
    \emph{and} direct-flow repair differ) and the \textsc{Func-2/4}
    $\to$\textsc{Func-10/10} pair (a scoring waiver changed) --- is
    labeled directional where it appears.
  \item \textbf{Cost accounting.} Token totals from resumed runs
    undercount, because resumed samples log zero tokens; within the
    cache trio the only clean full-run token cost is
    \textsc{Cache-off}'s 46.3\,M (the \textsc{Func} arms are
    uninterrupted full runs, so their totals are clean). Direct-flow
    repair calls are not instrumented, so direct-flow token totals ---
    including \textsc{Router-v1}'s 28.3\,M and \textsc{Router-v2}'s
    52.2\,M --- are lower bounds. No headline rests on them.
    Section~\ref{sec:results-routing} already sets aside the v1 arm's
    per-solved-task advantage as fitted to pre-audit evidence, and full
    instrumentation would only raise \textsc{Router-v2}'s total
    (its 110 direct samples are the only uninstrumented part), leaving
    the $\approx$all-planning-cost conclusion unchanged.
    Wall-clock is never compared across days (shared server, varying
    load).
  \item \textbf{Scoring strictness.} The harness fails on warnings, so
    our syntax rates are conservative relative to error-only benchmarks
    --- and Section~\ref{sec:results-artifacts} showed our own internal
    gate was at times \emph{stricter than the scorer}, deflating four
    tasks.
  \item \textbf{Benchmark fix boundary.} The AES S-box testbenches were
    repaired upstream on June~25,~2026, after our report, and the run
    inventory straddles that date. \textsc{Direct-t0}, \textsc{Pipe-t0},
    the cache trio, \textsc{Pipe-120b}, and \textsc{Func-2/4} were
    scored on the broken benches (their S-box failures carry the race's
    766-mismatch signature), while \textsc{Func-10/10},
    \textsc{Func-10/10-120b}, \textsc{Hybrid}, and every router arm
    postdate the fix.
    Comparisons whose arms sit on the same side are unaffected. Two
    comparisons span the boundary --- \textsc{Func-2/4}
    $\to$~\textsc{Func-10/10} (Section~\ref{sec:results-funcrepair})
    and \textsc{Func-10/10} vs.\ \textsc{Pipe-120b}
    (Section~\ref{sec:results-passk}) --- and both are therefore also
    reported with the two S-box tasks excluded; no other conclusion
    rests on those two tasks.
  \item \textbf{Ref-hack exclusion.} All rates exclude samples that
    instantiate a leaked \texttt{ref\_*} reference model
    (Section~\ref{sec:results-artifacts}). Detection is a syntactic
    screen over generated RTL; it flagged only unambiguous
    self-hacks and cannot, by construction, catch subtler uses of the
    leaked \emph{name} --- though the reference \emph{source} was never
    visible to the models, bounding what a subtler exploit could copy.
    The archived pre-fix runs retain the leak in the direct-flow
    prompt; \textsc{Router-v2} already runs with the filtered prompt
    (and produced zero hack samples), and re-running the remaining
    direct-flow arms with it is future work
    (Section~\ref{sec:future}).
  \item \textbf{Oracle feedback.} The functional repair loop consumes
    the scoring testbench's mismatch report. Its passes mean
    ``specification plus oracle feedback'' and are quarantined in
    Section~\ref{sec:results-funcrepair}; no headline number includes
    them. Relatedly, the internal lint compiles the testbench, so
    Verilator diagnostics can quote testbench lines into repair prompts;
    a flag exists to disable this for leak-free claims, at a syntax-rate
    cost we did not pay here.
  \item \textbf{Benchmark scope.} All results are on one benchmark's 60
    module-level tasks and three IP projects, with open-weight models of
    one family for all controlled comparisons. The family-level
    heterogeneity we observe (AES vs.\ E203) is itself evidence that
    results will vary across codebases.
\end{enumerate}

\section{Future Work}
\label{sec:future}

Five directions follow directly from the measured constraints.
\emph{First}, functional feedback without oracle leakage. Synthesize
self-checking benches or assertions from the specification, so the
functional repair loop gets a mismatch signal the scorer never sees ---
and gate it off pure-combinational modules, where
Section~\ref{sec:results-artifacts} showed the signal can be corrupted.
\emph{Second}, combine the pipeline with a testbench-generation
pipeline to unlock \emph{fine-grained module-decomposition IP reuse}.
Today the plan's decomposition stops at module granularity because only
the top module has an oracle. A companion pipeline that generates a
bench per planned submodule would let the system decompose more
aggressively, verify each decomposed piece independently, and reuse IP
at the sub-block level, turning the planner's decomposition from a
prompt-structuring device into a verified construction process.
\emph{Third}, interface-contract enforcement as a hard gate. The
testbench DUT instantiation is already injected as text; promoting it to
a mechanical post-generation port-diff check would remove the
\texttt{PINNOTFOUND} class that dominates giant-module syntax failures,
and the internal repair gate should mirror the scorer's warning waivers
exactly. \emph{Fourth}, selective decomposition for the giants.
RealBench's system-level tasks decompose into submodules that are
themselves benched module tasks, so a tree-structured generator could
verify each node against a \emph{real} testbench and compose upward.
This is decomposition with per-node oracles, reserved for the
scale-limited tier where it pays, and the natural first customer of the
testbench-generation direction above. \emph{Fifth}, hardening the
router. Recompute the oracle label table from cleaned evidence so
that confusion matrices are citable again; search for
planning-expensive tasks the direct flow genuinely solves --- the only
place a routing compute win can come from, a set that stays close to
empty at 120B once deep repair budgets are allowed
(Section~\ref{sec:results-routing}); extend the
\texttt{ref\_*}-filtered direct prompt --- already validated in the
\textsc{Router-v2} run, which produced zero hack samples --- to
re-runs of the archived direct-flow arms as the definitive control for
the ref-hack exclusion of Section~\ref{sec:results-artifacts}; and
replicate across seeds to make the routing result statistically
settled.

\section{Conclusion}
\label{sec:conclusion}

This thesis asked how far careful \emph{system} engineering --- planning,
grounding, verification, repair, and routing around fixed open-weight
models --- can push LLM-based RTL generation on a realistic,
reuse-dominated benchmark. The answer is a long way, and measurably so.
A two-stage pipeline that plans IP reuse against a grounded per-task
catalog and generates under environment-faithful verification reaches
$2.5\times$ the syntax rate and $3.0\times$ the pass rate of the same
model prompted directly, with the entire gain concentrated where reuse
exists to be exploited. A per-task router whose decision rule was re-derived on the audited
solution set keeps the fitted rule's coverage at 21 of 60 tasks
with higher per-sample pass and syntax rates and no exposure to the
benchmark's leaked reference models.

The thesis's second kind of contribution is measurement discipline over
the same system. Grounding made every plan correct and auditable.
Caching concentrated passing density inside solvable tasks. The repair
budgets were calibrated to where the rescues actually happen, recovering
the large majority of the functional phase's compute. A golden-model
audit of persistent zeros cleared the models on two tasks (and improved
the benchmark upstream), caught our own repair gate being stricter than
the scorer on three more, exposed a further benchmark defect on a
fourth, and confirmed the remaining zeros as genuine difficulty.
Turned on our own passes, the same audit caught a
prompt-side leak that had inflated the routing ablation, corrected it
before it could mislead, and led to the routing rule being re-derived
on the corrected evidence. These findings redirected the engineering
each time, and one was adopted upstream by the benchmark itself. The
frontier, at the end, is not syntax, not plans, and not scale. It is
functional correctness, and the next system to move it will be the one
that finds its models a better feedback signal than a mismatch count.
