% !TeX root = ../main.tex

% TODO(author): 中文摘要為英文摘要之翻譯初稿，請核對術語與行文後定稿。
\begin{abstract}

大型語言模型（LLM）已能生成語法正確的程式碼，然而在真實硬體專案中，暫存器傳輸級
（RTL）設計工作多以整合與矽智財（IP）重用為主，而非從零撰寫演算法模組。本論文以
RealBench 基準測試中源自三個真實開源 IP 專案（AES 加密核心、SD 卡控制器、E203
RISC-V 處理器核心）的 60 個模組級任務為評測對象，研究如何以系統工程手段——規劃、
接地（grounding）、驗證、修復與路由——提升固定開源權重模型的 RTL 生成能力。

本論文提出兩項主要貢獻。其一為兩階段代理式 IP 重用流水線：規劃代理透過
IP 重用工具（目錄檢索、候選 IP 檢視與評分、介面相容性檢查）針對每個任務
的真實 IP 目錄進行規劃，並以目錄注入、封閉詞彙表對齊、完整性檢查的三階段檢查
產生可稽核的結構化重用計畫，再由生成器在Verilator 驗證與修復迴圈下
產生 Verilog。在 T 為零的對照實驗中，流水線通過率達 25.0\%（語法 61.7\%），
為同一模型直接提示（8.3\%／25.0\%）的 3.0 倍，且增益完全集中於 IP 重用密集的
E203 家族。其二為逐任務路由器，依任務性質在完整流水線與低成本直接生成之間選擇流程。
本論文的基準測試作弊稽核發現，原始路由規則的極性係擬合於受污染的解答集；
改以清洗後證據重新推導規則——預設走流水線，僅窄小的自含運算葉模組
走直接生成——後，路由器維持全流水線基準臂 21/60 的解題數，單樣本通過率
與語法率均優於原規則，且完全不受洩漏影響；以清洗後樣本在較大模型規模下
重新評分時，更額外解出所有全流水線實驗臂皆未解出的一題。

本論文同時以對照消融精確量化各輔助機制：語意修復快取單調提升語法率
（59.4\%→66.9\%），並使通過樣本更集中於可解任務；兩種修復階段皆呈
「速修或永不修」分布，精簡的修復預算（語法約 6 次、功能不超過 4 次）
即可保留幾乎全部的修復成功案例，同時大幅節省運算；以黃金模組
檢驗持續零分任務，發現並向上游回報一項測試平台競爭條件（已獲基準採納
修復），亦確認其餘零分為真實難度；同一檢測反向檢驗通過樣本時，
更查獲一項提示端洩漏——直接生成提示中出現基準參考模型名稱，使部分樣本
僅靠實例化黃金模型即可通過——本文所有數據均已剔除此類樣本。
整體結果顯示：下一步的突破點不在
語法、規劃或模型規模，而在功能正確性——需要比失配計數更豐富的回饋
訊號。

\end{abstract}

\begin{abstract*}

Large language models generate syntactically plausible code, but real-world
RTL engineering is dominated by integration and IP reuse rather than
freestanding algorithm writing. This thesis studies how far system
engineering --- planning, grounding, verification, repair, and routing
around fixed open-weight models --- can push LLM-based Verilog generation
on RealBench, a benchmark of 60 module-level tasks drawn from three real
open-source IP projects (an AES cipher core, an SD-card controller, and
the E203 RISC-V processor core) and scored by a strict warnings-as-fatal
compile-and-simulate harness.

The thesis makes two primary contributions. First, a two-stage agentic
IP-reuse pipeline. A planning agent works each task through a suite of
IP-reuse tools --- catalog search, candidate inspection and scoring, and
interface-compatibility checking --- and produces a structured, auditable
reuse plan grounded by construction against a per-task catalog of the
modules the environment actually provides (catalog injection,
closed-vocabulary alignment, and a completeness gate). A generator then
works under a Verilator verify-and-repair loop that mirrors the scoring
environment exactly. At temperature zero the pipeline passes
25.0\% of tasks (61.7\% syntax) against the same model's direct-prompt
8.3\% (25.0\%), a $3.0\times$ improvement concentrated entirely in the
reuse-rich processor-core family. Second, a per-task router that
matches task character to generation flow between the full pipeline
and a cheap direct flow. The thesis's own benchmark-gaming audit
showed the rule's original polarity had been fitted to a contaminated
solution set, so the rule was re-derived on cleaned evidence ---
pipeline by default, direct generation only for narrow self-contained
computational leaves. The re-derived router holds the all-pipeline
arm's coverage (21 of 60 tasks solved) with higher per-sample pass and
syntax rates than the original rule and no exposure to the leak, and,
re-scored at the larger model scale, recovers one task that
all-pipeline arms miss.

Controlled ablations quantify the supporting machinery with equal care.
A semantic repair cache lifts syntax monotonically (59.4\%$\to$66.9\%)
while concentrating passing samples inside solvable tasks. Both repair
phases converge within their first attempts, so compact budgets
($\approx$6 syntax, $\leq$4 functional) keep essentially all rescues
at a fraction of the compute. A golden-model audit of persistently
failing tasks uncovered a testbench race condition --- which the
benchmark has since fixed upstream --- while confirming the remaining
zeros as genuine difficulty. Turned on the passes as well, the same audit
caught a prompt-side leak of reference-model names that had let
directly-prompted samples pass by wrapping the benchmark's own golden
model; every reported number excludes these samples. The clearest opportunity at the end is neither
syntax, nor planning, nor model scale, but functional correctness, which
will reward a richer feedback signal than a mismatch count.

\end{abstract*}
